Live Functional Programming with Typed Holes

### 2018-04-11 calculus.tex
small edits to sec 3

As with the type system for the external language (above), 
we specify expansion in the bidirectional style, via two mutually defined judgements.
%
The synthetic expansion judgement~$\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$ synthesizes a type~$\htau$ from~$\hexp$.
%
Synthetic expansion produces an internal expression~$d$ and a hole context~$\hDelta$.
%
%We say more about hole contexts, which are used in the type assignment judgement, $\hasType{\Delta}{\hGamma}{d}{\htau}$, below.
We discuss hole contexts further below, including their role in the type assignment judgement~$\hasType{\Delta}{\hGamma}{d}{\htau}$.

The analytic expansion judgement~$\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$, checks~$\hexp$ against~$\htau$ and produces an expansion~$d$ of type~$\htau'$, and a hole context~$\hDelta$.
%
The governing theorem below establishes that expansions are well-typed and in the analytic case that the type $\htau'$ is necessarily consistent with $\htau$.
%
% \rkc{minor, but the syntax of type assignment isn't mentioned before use. it's not
% entirely standard, because of the hole context, so perhaps worth mentioning.}
%
\begin{thm}[Typed Expansion]\label{thm:typed-expansion} ~
  \begin{enumerate}[nolistsep]
    \item
      If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      then $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
    \item
      If $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
      then $\tconsistent{\htau}{\htau'}$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$.
  \end{enumerate}
\end{thm}
\noindent

@@

The editor component of the implementation is derived from \Hazelnut. 
It exposes a language of structured edit actions (summarized in the sidebar on the left of Fig.~\ref{fig:screenshot}) that insert holes automatically to guarantee that every edit state has some (possibly incomplete) type. 
Formally,
this corresponds to the Sensibility property from the prior work \cite{popl-paper}. 
By composing this property with the Expandability, Typed Expansion and Progress properties from this paper, we establish a uniquely powerful 
invariant: that every edit state is both statically and dynamically meaningful. 
In other words, live feedback does not ``flicker in and out'' or reflect past edit states. 
The primary purpose of the included implementation is to serve as a proof-of-concept of this invariant, and to be of use to researchers studying the calculus as presented in this section.
% Consistent with this goal, it closely follows the theoretical account in this section, rather than including advanced language features.

>>>>

As with the type system for the external language (above), 
we specify expansion bidirectionally \cite{DBLP:conf/ppdp/FerreiraP14}.
%
The synthetic expansion judgement~$\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$ synthesizes a type~$\htau$ from~$\hexp$.
%
Synthetic expansion produces an internal expression~$d$ and a hole context~$\hDelta$.
%
%We say more about hole contexts, which are used in the type assignment judgement, $\hasType{\Delta}{\hGamma}{d}{\htau}$, below.
We describe hole contexts, which serve as ``inputs'' to the type assignment judgement~$\hasType{\Delta}{\hGamma}{d}{\htau}$, further below. 
%
The analytic expansion judgement~$\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$, checks~$\hexp$ against~$\htau$ and produces an expansion~$d$ of type~$\htau'$, and a hole context~$\hDelta$.
%
The governing theorem below establishes that expansions are well-typed and in the analytic case that the type $\htau'$ is consistent with $\htau$.
%
\begin{thm}[Typed Expansion]\label{thm:typed-expansion} ~
  \begin{enumerate}[nolistsep]
    \item
      If $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      then $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
    \item
      If $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
      then $\tconsistent{\htau}{\htau'}$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau'}$.
  \end{enumerate}
\end{thm}
\noindent

@@

The editor component of the implementation is derived from \Hazelnut. 
It exposes a language of structured edit actions (summarized in the sidebar on the left of Fig.~\ref{fig:screenshot}) that insert holes automatically to guarantee that every edit state has some (possibly incomplete) type. 
Formally,
this corresponds to the Sensibility property from the prior work \cite{popl-paper}.\todo{put statement in the appendix}{} 
By composing this property with the Expandability, Typed Expansion and Progress properties from this paper, we establish a uniquely powerful 
invariant: that every edit state is both statically and dynamically meaningful. 
In other words, live feedback does not ``flicker in and out'' or reflect past edit states. 
The primary purpose of the included implementation is to serve as a proof-of-concept of this invariant, and to be of use to researchers studying the calculus as presented in this section.
% Consistent with this goal, it closely follows the theoretical account in this section, rather than including advanced language features.
